Dark matter halo shape at $z=0$ in the Auriga simulations: radial evolution and alignment with the stellar disk

### 2019-07-18 Document/paper.tex
documento

A robust prediction of the Cold Dark Matter (CDM) paradigm is that DM
halos are ellipsoidal and can be characterized by the principal axes
$a>b>c$.
This ellipsoidal shape is mostly due to the anisotropical and
clumpy accretion of matter influenced by environmental structures.
Numerical studies how that the shape has a strong mass dependence
\citep{Allgood_et_al._2006}, halos are also rounder at the outerskirts
than at the inner part. 
Shape also evolves with cosmic time, halos get
rounder as they evolve.

>>>>

Accurate observations and modeling of our Galaxy can also
shape our understainding of the Universe as a whole.  
Explaining its current composition and kinematical state is 
equivalent to figuring out its formation process in a cosmological
context. 
As a first approximation, the Milky Way's (MW) morphology and
separation into global kinematically coherent components, such as its
disk and bulge, can be used to support the  existence of a Dark Matter
(DM) component to explain its dynamics. 
In this context the most detailed observations and modelling has been
devoted to the region around the solar neighborhood
\citep{2000MNRAS.311..361O,2009PASJ...61..227S,2010JCAP...08..004C,2013ApJ...779..115B,Iocco15}.

Improvements in the sensitivity of astronomical observations have 
extended this kind of modelling to outter regions by measuring
the stellar streams resulting from infalling globular clusters or
satellite galaxies that got tidaly disrutpted by the gravitational
potential of the Milky Way
\citep{1998ApJ...495..297J,1999MNRAS.307..495H, 1999MNRAS.307..877T}. 
The interpretation of such fossil records requires the the
determination of the full three-dimensional gravitational potential of
the Milky  Way.

The detailed constraints on the shape of this potential can be then
confronted against the expectations from different galaxy formation
models in an explicit cosmological context.
For instance, in the current dominant paradigm of Cold Dark Matter
(CDM) dominated Universe galaxies are expected to be hosted by
triaxial DM halos.

This ellipsoidal shape is mostly due to the anisotropical and
clumpy accretion of matter influenced by environmental structures.
Numerical studies how that the shape has a strong mass dependence
\citep{Allgood_et_al._2006}, halos are also rounder at the outerskirts
than at the inner part. 
Shape also evolves with cosmic time, halos get
rounder as they evolve.
